### 2012-11-07 paper/isca40format/AbstractProtocol.tex
with trigger

\newcommand{\One}[2]{\text{#1$\langle#2\rangle$}}
\newcommand{\two}[3]{\text{#1$\langle#2,#3\rangle$}}
\newcommand{\three}[4]{\text{#1$\langle#2,#3,#4\rangle$}}
\newcommand{\four}[5]{\text{#1$\langle#2,#3,#4,#5\rangle$}}
\newcommand{\True}{\texttt{True}}
\newcommand{\False}{\texttt{False}}
\newcommand{\assigns}{:=}
\newcommand{\AndAnd}{\;\&\&\;}
\newcommand{\OrOr}{\;||\;}

@@

The notation we adopt is primarily for conciseness of description while allowing
maximum flexibility in the implementation. An realistic efficient implementation
would forego some of the flexibility, and we will show such implementations
later.

@@

\begin{figure}
\begin{tabular}{|l|p{.64\columnwidth}|}
\hline
isPresent(a) & Cache line for address $a$ is present\\
\hline
data(a) & Cache line data for address $a$\\
\hline
state(a) & Coherence state for address $a$\\
\hline
dir(a, c) & Directory's copy of the coherence state for address $a$ in child
$c$\\
\hline
isCWait(c, a) & Check if the node waiting for a response from child $c$
for address $a$\\
\hline
cWait(cm a) & The coherence state to which the node has requested child $c$ to
go to, for address $a$\\
\hline
isPWait(a) & Check if the node waiting for a response from its parent for
address $a$\\
\hline
replacingAddr(a) & The address of the line that is going to replace the line
corresponding to address $a$\\
\hline
\end{tabular}
\caption{Information that the node keeps for a cache line}
\label{table:states}
\end{figure}

Figure~\ref{table:messages} shows the incoming and outgoing messages to and from
the node, respectively.

@@

Each message coming from a particular child or the parent is independent of the
messages coming from another source, in addition to being independent of the
messages for a different address. The node can process messages coming from
different sources, and for different addresses in any order. \emph{ A response
or a request should not be handled before handling an earlier response for the
same address from the same source}. Even though our abstract protocol requires
this ordering, we later show that is easy to enforce these even if the
underlying network doesn't enforce point-to-point FIFO ordering or priorities
between different kinds of messages.

Evicting a cache line for replacement can also be treated as a request message,
albeit generated internally by the node. Such a request would be generated if a
cache line requested by a child is currently not present in the cache, and some
other cache line has to be evicted to make room for the requested cache line.

To allow maximum flexibility of implementation, the node is given separate
access queues for each kind of message (request of response) from each source
and for each address. An implementation of a protocol using our methodology does
not have to keep a separate access queue for each source and each address, but
it must adhere to the following rules: \emph{\emph{a)} a request from any child
should not block a request from the parent, and \emph{b)} no request should
block any response}. As we show later, these rules are necessary to avoid
deadlocks. This flexibility allows one to design a wide range of blocking
policy.

The values of the coherence state (and hence the directory state) belong to a
totally ordered set, \ie a set in which a \emph{less than} ($<$) relation is
defined between every pair of elements in that set. If a line is not present,
then the state of the line is the least value in this totally ordered set, for
comparison purposes. Let us call this least value as $Inv$, a shorthand for
Invalid. Typically, a line in state $Inv$ signifies no permissions.

We now describe the procedures for a node to handle requests and responses. Note
that a message handler would be triggered only if the corresponding message is
present.  Similarly, if the required resources for handling a message are not
present (for example, if the handling of the request from a child requires
sending further requests to its children or the parent, and the send buffers are
full), then the handler for request from a child would not be triggered.

Figure~\ref{alg:handlePReq} specifies how to handle a node's incoming request
from the parent for downgrading the coherence state of address $a$ to $x$.

Function $hasDataXfer(x, y)$ specifies whether data has to be transferred
from a node to its parent when the node downgrades the coherence state of some
line from $x$ to $y$.  This function is used at the node to determine if the
child has transferred data or not on receiving a response from the child, and
also to determine if the node has to send data to the parent, for processing a
request from the parent. In a typical coherence protocol, the data has to be
only transferred by a node to its parent if the cache line has been modified.

@@

\begin{figure}
\begin{algorithmic}
\Function{handlePReq}{\two{ReqFromP}{a}{x}}
\If {$!isPresent(a)$}
  \State remove$(\two{ReqFromP}{a}{x})$
\ElsIf {$\forall c, !isCWait(c, a)$}
  \If {$state(a) \le x$}
     \State remove$(\two{ReqFromP}{a}{x})$
  \ElsIf {$\forall c, dir(c, a) \le x$}
     \State remove$(\two{ReqFromP}{a}{x})$
     \If {hasDataXfer$(state(a), x)$}
        \State send$(\two{DataToP}{a}{data(a)})$
     \EndIf
     \State $state(a) \assigns x$
     \State send$(\two{RespToP}{a}{x})$
  \Else
     \ForAll{$c$}
        \If {$dir(c, a) > x$}
           \State $isCWait(c, a) \assigns \True$
           \State $cWait(c, a) \assigns x$
           \State send$(\three{ReqToC}{c}{a}{x})$
        \EndIf
     \EndFor
  \EndIf
\EndIf
\EndFunction
\end{algorithmic}
\caption{Handle request from parent}
\label{alg:handlePReq}
\end{figure}

@@

\begin{figure}
\begin{algorithmic}
\Function{handlePResp}{\two{RespFromP}{a}{x}}
  \State remove$(\two{RespFromP}{a}{x})$
  \If {$state(a) \neq Inv$}
     \State remove$(\two{DataFromP}{a}{d})$
     \State $data(a) = d$
  \EndIf
  \State $state(a) \assigns x$
  \State $isPWait(a) \assigns \False$ 
\EndFunction
\end{algorithmic}
\caption{Handle response from parent}
\label{alg:handlePResp}
\end{figure}

@@

\begin{figure}
\begin{algorithmic}
\Function{handleCResp}{\three{RespFromC}{c}{a}{x}}
  \State remove$(\two{RespFromC}{c}{a}{x})$
  \If {hasDataXfer$(dir(c,a), x)$}
     \State remove$(\two{DataFromC}{c}{a}{d})$
     \State $data(a) = d$
  \EndIf
  \State $dir(c,a) \assigns x$
  \If {$isCWait(c, a) \AndAnd cWait(c, a) >= x$}
    \State $isCWait(c, a) \assigns \False$
  \EndIf
\EndFunction
\end{algorithmic}
\caption{Handle response from child}
\label{alg:handleCResp}
\end{figure}

Figure~\ref{alg:handleCReq} specifies how to handle a node's incoming request
from a child for upgrading the child's coherence state for address $a$ to $x$.

Function \emph{compatible(x)} gives the highest state that a line in another
cache can be in without violating coherency if the line is in one cache in state
$x$. For example, the highest state that is compatible with a state having
read-write permissions is the state having no permissions (\ie $Inv$).

Function \emph{getReplace(a)} gives the address of the line that can be replaced
in order to accommodate the line in address $a$. However, if a line is waiting
for a response (from either a child or the parent), then that line can not be
replaced (ensured by the check for $isCWait$ and $isPWait$) in the pseudo-code.

\begin{figure}
\begin{algorithmic}
\Function{handleCReq}{\three{ReqFromC}{c}{a}{x}}
  \If {$isPresent(a)$}
     \If {$!isPWait(a)\ \AndAnd \forall c, !isCWait(c, a)$}
        \If {$state(a) \ge x \AndAnd$ \\
              \hspace{2.5cm} $\forall c' \neq c, dir(c', a) \le \text{compatible}(x)$}
           \State remove$(\three{ReqFromC}{c}{a}{x})$
           \If {$dir(c,a) \neq Inv)$}
              \State send$(\three{DataToC}{c}{a}{data(a)})$
           \EndIf
           \State $dir(c, a) \assigns x$
           \State send$(\three{RespToC}{c}{a}{x})$
        \Else
           \If {$state(a) < x$}
              \State $isPWait(a) \assigns \True$
              \State send$(\two{ReqToP}{a}{x})$
           \EndIf
           \ForAll {$c' \neq c$}
              \If {$dir(c', a) > \text{compatible}(x)$}
                 \State $isCWait(c', a) \assigns \True$
                 \State $cWait(c', a) \assigns \text{compatible}(x)$
                 \State send$(\three{ReqToC}{c'}{a}{\text{compatible}(x)})$
              \EndIf
           \EndFor
        \EndIf
     \EndIf
  \Else
     \State \textbf{let} $a' = \text{getReplace}(a)$
     \If {$!isPWait(a') \AndAnd !isCWait(a')$}
        \State $replacingAddr(a') \assigns a$
        \State send$(\One{EvictReq}{a'})$
     \EndIf
  \EndIf
\EndFunction
\end{algorithmic}
\caption{Handle request from child}
\label{alg:handleCReq}
\end{figure}

Figure~\ref{alg:handleEvict} specifies how to handle an eviction message (in
order to replace the current line in address $a$ with the line for some other
address, which is given by $replacingAddr(a)$).
\begin{figure}
\begin{algorithmic}
\Function{handleEvict}{\One{EvictReq}{a}}
   \If {$!isPresent(a) \OrOr$\\
        \hspace{3cm} $\forall c, dir(c, a) = Inv$}
      \State remove$(\One{EvictReq}{a})$
      \If {$state(a) \neq Inv$}
         \State send$(\two{RespToP}{a}{Inv})$
         \If {hasDataXfer$(state(a), Inv)$}
            \State send$(\two{RespToP}{a}{data(a)})$
         \EndIf
      \EndIf
      \State $isPresent(a) \assigns \False$
      \State $isPresent(replacingAddr(a)) \assigns \True$
   \Else
      \If {$\forall c, !isCWait(c, a)$}
         \ForAll{$c$}
            \If {$dir(c, a) > Inv$}
               \State $isCWait(c, a) \assigns \True$
               \State $cWait(c, a) \assigns Inv$
               \State send$(\three{ReqToC}{c}{a}{Inv})$
            \EndIf
         \EndFor
      \EndIf
   \EndIf
\EndFunction
\end{algorithmic}
\caption{Evict a line in address $a$}
\label{alg:handleEvict}
\end{figure}

@@

Note that the handlers continue to be triggered till the corresponding message
has been removed.

In order to avoid starvation, once a response is available, the handler for that
response must be triggered eventually.

Once a node starts handling a request from the parent, it can ignore all other
requests (from the children, or eviction requests) to the same address, till the
node has removed the request from the parent.

Similarly, once a node starts handling a request from a child, it can ignore
other requests from other children to the same address, till it has removed the
request from the child. However, it can not ignore requests from the parent to
the same address, in order to avoid deadlocks.

>>>>

\newcommand{\One}[2]{\text{#1$\langle#2\rangle$}}
\newcommand{\two}[3]{\text{#1$\langle#2,#3\rangle$}}
\newcommand{\three}[4]{\text{#1$\langle#2,#3,#4\rangle$}}
\newcommand{\four}[5]{\text{#1$\langle#2,#3,#4,#5\rangle$}}
\newcommand{\True}{\texttt{True}}
\newcommand{\False}{\texttt{False}}
\newcommand{\assigns}{:=}
\newcommand{\AndAnd}{\;\&\&\;}
\newcommand{\OrOr}{\;||\;}
\newcommand{\Inv}{\text{$Inv$}}

\definecolor{trig-color}{gray}{.6}

@@

The notation we adopt is primarily for conciseness of description while allowing
maximum flexibility in the implementation.

@@

\begin{figure}
\begin{tabular}{|l|p{.64\columnwidth}|}
\hline
isPresent(a) & Check if cache line for address $a$ is present\\
\hline
data(a) & Cache line data for address $a$\\
\hline
state(a) & Coherence state for address $a$\\
\hline
dir(a, c) & Directory's copy of the coherence state for address $a$ in child
$c$\\
\hline
isCWait(c, a) & Check if the node waiting for a response from child $c$
for address $a$\\
\hline
cWait(c, a) & The coherence state to which the node has requested child $c$ to
go to, for address $a$\\
\hline
isPWait(a) & Check if the node waiting for a response from its parent for
address $a$\\
\hline
replacingAddr(a) & The address of the line that is going to replace the line
corresponding to address $a$\\
\hline
whoTrigger(a) & The source of the request which made the node wait for a
response. It is $None, Parent, Child \text{ or } Evict$\\
\hline
childTrigger(a) & If $whoTrigger(a)$ is $Child$, the index of the child which
is the source of the request\\
\hline
\end{tabular}
\caption{Information that the node keeps for a cache line}
\label{table:states}
\end{figure}

The values of the coherence state (and hence the directory state) belong to a
totally ordered set, \ie a set in which a \emph{less than} ($<$) relation is
defined between every pair of elements in that set and that relation is
transitive. For instance, in an MSI protocol, the values of the coherence state
are from the set $\{M, S, I\}$ where $I < S$, $S < M$ and $I < M$. Our abstract
protocol does not assign any semantics or meaning to the values of the coherence
states. In a concrete cache coherence protocol, the values of the states are
associated with permissions for reading and writing, and the state with fewer
permissions is $<$ the state with more permissions. We use \Inv to denote the
least element in the set containing the values of the coherence state.
Typically, this denotes that the node has no permissions for reading or writing
that line.

In order to increase the coherence state of a cache line, a node has to send an
\emph{upgrade request} to its parent and may eventually get an \emph{upgrade
response} from the parent. In order to decrease the coherence state of a cache
line for one of the node's children, the node has to send a \emph{downgrade
request} to that child, and may eventually get a \emph{downgrade response} from
that child.

Figure~\ref{table:messages} shows the incoming and outgoing messages to and from
the node, respectively, and requests for eviction. An eviction request for an
address is generated if a child has made a request for a different cache line
whose address is not currently present in the cache.

@@

@@

\begin{figure}
\begin{minipage}{\columnwidth}
\begin{enumerate}
\item A request or a response should not be handled before handling an earlier
response for the same address from the same source

\item A request from a child for an address should not block or prevent the
handling of a request from the parent for the same address

\item A request for an address should not block or prevent the handling of a
response from any source for any address
\end{enumerate}
\end{minipage}
\caption{Message ordering, and blocking restrictions}
\label{list:messageOrder}
\end{figure}

Each message coming from a particular child or the parent is independent of the
messages coming from another source, in addition to being independent of the
messages for a different address. The node can process messages coming from
different sources, and for different addresses in any order, subject to the
ordering requirements in Figure~\ref{list:messageOrder}. We later show that is
easy to enforce this ordering even if the underlying network doesn't enforce
point-to-point FIFO ordering or priorities between different kinds of messages.

To allow maximum flexibility of implementation, the node is given separate
access queues for each kind of message (request of response) from each source
and for each address. An implementation of a protocol using our methodology does
not have to keep a separate access queue for each source and each address, but
it must adhere to the blocking restrictions in Figure~\ref{list:messageOrder}.
These restrictions are necessary to avoid deadlocks. The flexibility of the
abstract protocol allows one to design a wide range of blocking policies.

Our abstract protocol also depends on the designer correctly defining the three
functions shown in Figure~\ref{table:functions}. These functions abstract away
the coherence aspect of the cache coherence protocol. Note that $maxCompat$
function can be derived from $isCompat$ function, if $isCompat$ is defined
correctly. Figure~\ref{fig:msiFunc} shows the functions defined for MSI (--
denotes undefined).

\begin{figure}
\begin{tabular}{|c|p{.63\columnwidth}|}
\hline
isCompat(x, y) & Checks if two different caches can have coherence states
$x$ and $y$, respectively, for the same address\\
\hline
maxCompat(x) & Returns a coherence state value $y$ such that, $\forall z \le y,
isCompat(x, z)$ and $\forall w > y, !isCompat(x, w)$\\
\hline
dataXfer(x, y) & Checks if cache line data has to be transferred by a cache to its
parent on downgrading from $x$ to $y$\\
\hline
\end{tabular}
\caption{Functions to abstract away the coherence aspect}
\label{table:functions}
\end{figure}

\begin{figure}
\centering
\begin{subfigure}{.45\columnwidth}
\centering
\begin{tabular}{|c||c|c|c|}
\hline
& I & S & M\\
\hline\hline
I & $\checkmark$ & $\checkmark$ & $\checkmark$\\
\hline
S & $\checkmark$ & $\checkmark$ & $\times$\\
\hline
M & $\checkmark$ & $\times$ & $\times$\\
\hline
\end{tabular}
\label{table:isCompat}
\caption{$isCompat$ for MSI}
\end{subfigure}
\quad
\begin{subfigure}{.45\columnwidth}
\centering
\begin{tabular}{|c||c|c|c|}
\hline
\backslashbox{$x$}{$y$} & I & S & M\\
\hline\hline
I & -- & -- & --\\
\hline
S & $\times$ & -- & --\\
\hline
M & $\checkmark$ & $\checkmark$ & --\\
\hline
\end{tabular}
\label{table:dataXfer}
\caption{$dataXfer(x, y)$ for MSI}
\end{subfigure}
\caption{Coherence abstraction for MSI}
\label{fig:msiFunc}
\end{figure}

We now describe the procedures for a node to handle requests and responses. Note
that a message handler would be triggered only if the corresponding message is
present. If the required resources for handling a message are not present (for
example, if the handling of the request from a child requires sending further
requests to its children or the parent, and the send buffers are full), then the
handler for request from a child would not be triggered.

Figure~\ref{alg:handlePReq} specifies how to handle a node's incoming request
from the parent for downgrading the coherence state of address $a$ to $x$. Note
that a request from the parent can be handled successfully only if the node is
not already waiting for a response from a child.

Figure~\ref{alg:handleCReq} specifies how to handle a node's incoming request
from a child for upgrading the child's coherence state for address $a$ to $x$.
Note that a request from a child can be handled successfully only if the node is
not already waiting for a response from the parent or from any child.

@@

@@

Figure~\ref{alg:handleEvict} specifies how to handle an eviction message (in
order to replace the current line in address $a$ with the line for some other
address, which is given by $replacingAddr(a)$).

\begin{figure}
\begin{algorithmic}
\Rule{handlePReq}{\two{ReqFromP}{a}{x}}
\If {$!isPresent(a)$}
  \State remove$(\two{ReqFromP}{a}{x})$
\ElsIf {$\forall c, !isCWait(c, a)$}
  \If {$state(a) \le x$}
     \State remove$(\two{ReqFromP}{a}{x})$
  \ElsIf {$\forall c, dir(c, a) \le x$}
     \State remove$(\two{ReqFromP}{a}{x})$
     \If {dataXfer$(state(a), x)$}
        \State send$(\two{DataToP}{a}{data(a)})$
     \EndIf
     \State $state(a) \assigns x$
     \State send$(\two{RespToP}{a}{x})$
  \Else
     \ForAll{$c$}
        \If {$dir(c, a) > x$}
           \State $isCWait(c, a) \assigns \True$
           \State $cWait(c, a) \assigns x$
           \State send$(\three{ReqToC}{c}{a}{x})$
        \EndIf
     \EndFor
     \State \textcolor{trig-color}{$whoTrigger(a) \assigns Parent$}
  \EndIf
\EndIf
\EndRule
\end{algorithmic}
\caption{Handle request from parent}
\label{alg:handlePReq}
\end{figure}

\begin{figure}
\begin{algorithmic}
\Rule{handleCReq}{\three{ReqFromC}{c}{a}{x}}
  \If {$isPresent(a)$}
     \If {$!isPWait(a)\ \AndAnd \forall c, !isCWait(c, a)$}
        \If {$state(a) \ge x \AndAnd$ \\
              \hspace{2.5cm} $\forall c' \neq c, dir(c', a) \le \text{maxCompat}(x)$}
           \State remove$(\three{ReqFromC}{c}{a}{x})$
           \If {$dir(c,a) \neq \Inv)$}
              \State send$(\three{DataToC}{c}{a}{data(a)})$
           \EndIf
           \State $dir(c, a) \assigns x$
           \State send$(\three{RespToC}{c}{a}{x})$
        \Else
           \If {$state(a) < x$}
              \State $isPWait(a) \assigns \True$
              \State send$(\two{ReqToP}{a}{x})$
           \EndIf
           \ForAll {$c' \neq c$}
              \If {$dir(c', a) > \text{maxCompat}(x)$}
                 \State $isCWait(c', a) \assigns \True$
                 \State $cWait(c', a) \assigns \text{maxCompat}(x)$
                 \State send$(\three{ReqToC}{c'}{a}{\text{maxCompat}(x)})$
              \EndIf
           \EndFor
           \State \textcolor{trig-color}{$whoTrigger(a) \assigns Child$}
           \State \textcolor{trig-color}{$childTrigger(a) \assigns c$}
        \EndIf
     \EndIf
  \Else
     \State \textbf{let} $a' \assigns \text{getReplace}(a)$
     \If {$!isPWait(a') \AndAnd !isCWait(a')$}
        \State $replacingAddr(a') \assigns a$
        \State send$(\One{EvictReq}{a'})$
        \State \textcolor{trig-color}{$whoTrigger(a) \assigns Evict$}
        \State \textcolor{trig-color}{$childTrigger(a) \assigns c$}
     \EndIf
  \EndIf
\EndRule
\end{algorithmic}
\caption{Handle request from child}
\label{alg:handleCReq}
\end{figure}

\begin{figure}
\begin{algorithmic}
\Rule{handleEvict}{\One{EvictReq}{a}}
   \If {$!isPresent(a) \OrOr \forall c, dir(c, a) = \Inv$}
      \State remove$(\One{EvictReq}{a})$
      \If {$state(a) \neq \Inv$}
         \State send$(\two{RespToP}{a}{\Inv})$
         \If {dataXfer$(state(a), \Inv)$}
            \State send$(\two{RespToP}{a}{data(a)})$
         \EndIf
      \EndIf
      \State $isPresent(a) \assigns \False$
      \State allocLine$(replacingAddr(a))$
      \State \textcolor{trig-color}{\textbf{trigger} handleCReq($childTrigger(a)$)}
   \Else
      \If {$\forall c, !isCWait(c, a)$}
         \ForAll{$c$}
            \If {$dir(c, a) > \Inv$}
               \State $isCWait(c, a) \assigns \True$
               \State $cWait(c, a) \assigns \Inv$
               \State send$(\three{ReqToC}{c}{a}{\Inv})$
            \EndIf
         \EndFor
      \EndIf
   \EndIf
\EndRule
\end{algorithmic}
\caption{Evict a line in address $a$}
\label{alg:handleEvict}
\end{figure}

\begin{figure}
\begin{algorithmic}
\Rule{handlePResp}{\two{RespFromP}{a}{x}}
  \State remove$(\two{RespFromP}{a}{x})$
  \If {$state(a) \neq \Inv$}
     \State remove$(\two{DataFromP}{a}{d})$
     \State $data(a) = d$
  \EndIf
  \State $state(a) \assigns x$
  \State $isPWait(a) \assigns \False$ 
  \State \textcolor{trig-color}{\textbf{trigger} handleCReq($childTrigger(a)$)}
\EndRule
\end{algorithmic}
\caption{Handle response from parent}
\label{alg:handlePResp}
\end{figure}

\begin{figure}
\begin{algorithmic}
\Rule{handleCResp}{\three{RespFromC}{c}{a}{x}}
  \State remove$(\two{RespFromC}{c}{a}{x})$
  \If {dataXfer$(dir(c,a), x)$}
     \State remove$(\two{DataFromC}{c}{a}{d})$
     \State $data(a) = d$
  \EndIf
  \State $dir(c,a) \assigns x$
  \If {$isCWait(c, a) \AndAnd cWait(c, a) >= x$}
    \State $isCWait(c, a) \assigns \False$
  \EndIf
  {\color{trig-color}
  \If {$whoTrigger(a) = Parent$}
     \State \textbf{trigger} handlePReq
  \ElsIf {$whoTrigger(a) = Evict$}
     \State \textbf{trigger} handleEvict
  \ElsIf {$whoTrigger(a) = Child$}
     \State \textbf{trigger} handleCReq($childTrigger(a)$)
  \EndIf}
\EndRule
\end{algorithmic}
\caption{Handle response from child}
\label{alg:handleCResp}
\end{figure}

Function \emph{getReplace(a)} gives the address of the line that can be replaced
in order to accommodate the line in address $a$. Note that if the line selected
for eviction is waiting for a response from the parent or from any child, then
it can not be replaced.

Figure~\ref{alg:allocLine} shows the initiation that has to happen for
allocating a new cache line for an address.

\begin{figure}
\begin{algorithmic}
\Procedure{allocLine}{$a$}
  \State $isPresent(a) \assigns \True$
  \State \textcolor{trig-color}{$whoTrigger(a) \assigns None$}
  \State $state(a) \assigns \Inv$
  \ForAll {$c$}
     \State $dir(c, a) \assigns \Inv$
  \EndFor
  \State $isCWait(a) \assigns \False$
  \State $isPWait(a) \assigns \False$
\EndProcedure
\end{algorithmic}
\caption{Allocate a new cache line for address $a$}
\label{alg:allocLine}
\end{figure}

@@

Each handler will be triggered repeatedly till the corresponding message has
been removed. In order to avoid starvation, the following two conditions must be
maintained:
\begin{enumerate}
\item Once a response is available from any source, the handler for that
response must be triggered eventually.
\item After handling a response (or an evict request) the corresponding handler
must trigger the original handler that requested the received response (or made
an eviction request). The $whoTrigger(a)$ keeps track of the original handler
for each address $a$. In case of a $Child$ trigger, the actual child
corresponding to the original handler is kept track of in $childTrigger(a)$, for
address $a$, in which case $handleCReq$ is triggered for the outstanding request
from that child. In the figures, the code corresponding to this is grayed out
\end{enumerate}

### 2012-11-07 paper/isca40format/Introduction.tex
with trigger

Large shared-memory multicores are likely to use cache-coherence protocols
which are directory based, hierarchical, and implemented using message passing.
Even though such protocols have been around for two decades (see, for example,
DASH\cite{DASH}, Alewife\cite{Anant}, ...) correct design and implementation of
such protocols remains a challenge. It is difficult to convince oneself that
all corner cases have been handled adequately without a formal proof. And yet
formal proofs of real implementations are practically non-existent. One can not
settle for conservative choices in cache-coherence protocol designs because
performance (\eg the number of messages needed to maintain coherence, the
buffer requirements, etc) is as important as correctness. Consequently proving
a simple protocol for a small number of caches, under unbounded buffer
assumption is of limited use in gaining confidence in the correctness of a real
protocol.

@@

\item For each message type, determining which cache and directory states are
possible and therefore must be dealt with is hard. An incomplete set of
transition rules may cause the protocol to deadlock.

@@

These examples give a glimpse of the complexity of designing and implementing a
correct cache coherence protocol in a message-passing directory-based setting.
Even for a 2-level cache hierarchy, designing and implementing the protocol is
very hard. This problem is greatly exacerbated in a multi-level cache
hierarchy, which is becoming the norm because of the increasing core counts.

@@

Our method substantially reduces the design and proof burden for new protocols
because of the common abstract protocol. The abstract protocol is a set of
transitions that the caches and directories follow to send or receive a request
or a response. We have formally proved that these transitions ensure the
following properties: a) the directory's notion of a cache's state is
``conservative'' in that a cache is never going to have more permissions that
what the directory thinks the cache has; and b) any request will eventually get
a corresponding response. A formal proof of the abstract protocol was written
using the Coq theorem prover \cite{}, however, a discussion of the formal proof
is beyond the scope of this paper.

@@

The main contributions of this paper are: 1. A new abstract protocol that
satisfies two important invariants for distributed and hierarchal
message-passing protocols; 2. Instantiation of the abstract protocol to
implement four well-known protocols (MSI, MESI, MOSI, MOESI) and their
variations and showing why their correctness is guaranteed as long as the
abstract protocol is correct, \ie it provably implements the two invariants;
and 3. Our method does not introduce any overhead over direct implementation of
these protocols.  We think that this new method of designing protocols
substantially reduces the required effort, and that our abstract protocol is
flexible enough to implement (while ensuring correctness of) most if not all
inclusive message-passing directory-based protocols.

>>>>

Large shared-memory multicores are likely to use cache-coherence protocols
which are directory-based, hierarchical, and implemented using message passing.
Even though such protocols have been around for two decades (see, for example,
DASH\cite{DASH}, Alewife\cite{Anant}, ...) correct design and implementation of
such protocols remains a challenge. It is difficult to convince oneself that
all corner cases have been handled adequately without a formal proof. And yet
formal proofs of real implementations are practically non-existent. One can't
settle for conservative choices in cache-coherence protocol designs because
performance (\eg the number of messages needed to maintain coherence, the
buffer requirements, etc) is as important as correctness. Consequently proving
a simple protocol for a small number of caches, under unbounded buffer
assumption is of limited use in gaining confidence in the correctness of a real
protocol.

@@

\item For each message type, determining which cache and directory states are
possible (and therefore must be dealt) with is difficult. An incomplete set of
transition rules may cause the protocol to deadlock.

@@

These examples give a glimpse of the complexity of designing and implementing a
correct cache coherence protocol in a message-passing, directory-based setting.
Even for a 2-level cache hierarchy, designing and implementing the protocol is
quite difficult. This problem is greatly exacerbated in a multi-level cache
hierarchy, which is becoming the norm because of the increasing core counts.

@@

Our method substantially reduces the design and proof burden for new protocols
because of the common abstract protocol. The abstract protocol is a set of
transitions that the caches and directories follow to send or receive a request
or a response. We have formally proved that these transitions ensure the
following properties: a) the directory's notion of a cache's state is
``conservative'' in that a cache is never going to have more permissions that
what the directory thinks the cache has; and b) any request will eventually get
a corresponding response. A formal proof of the abstract protocol was written
using the Coq theorem prover \cite{}, however, a presentation of the formal
proof is beyond the scope of this paper.

@@

The main contributions of this paper are: 1. A new abstract protocol that
satisfies two important invariants for distributed and hierarchal
message-passing protocols; 2. Instantiation of the abstract protocol to
implement four well-known protocols (MSI, MESI, MOSI, MOESI) and their
variations and showing why their correctness is guaranteed as long as the
abstract protocol is correct, \ie it provably implements the two invariants;
and 3. Our method does not introduce any overhead over direct implementation of
these protocols.  We think that this new method of designing protocols
substantially reduces the required effort, and that our abstract protocol is
flexible enough to implement (while ensuring correctness of) most if not all
inclusive, message-passing, directory-based protocols.
